---
title: "Homework - Week 01"
author: "Isabella C. Richmond"
format: html
---

```{r}
#| label: setup
#| echo: FALSE
#| eval: TRUE
#| warning: FALSE

library(rethinking)
library(ggplot2)
```

#### Question 1: Suppose the globe tossing data (Lecture 2, Chapter 2) had turned out to be 4 water and 11 land. Construct the posterior distribution.

```{r}
#| label: q1
#| echo: TRUE
#| eval: TRUE

# define how many tosses were W & L
w <- 4
l <- 11

# define p, every 0.01 between 0,1 
p <- seq(0, 1, by = 0.01)

# function to compute posterior by every p
compute_posterior <- function(W, L, poss){
  ways <- sapply(poss, function(q) (q*4)^W * ((1-q)*4)^L)
  post <- ways/sum(ways)
  data.frame(poss, ways, post = round(post,3))
}

# compute posterior for our data
posterior <- compute_posterior(w, l, p)

# visualize posterior 
ggplot(data = posterior) + 
  geom_line(aes(poss, post)) + 
  theme_classic() + 
  xlab("proportion water") + 
  ylab("probability")


```
NOTE: Bonus answer in solutions is somewhat key to understanding next answer. This specific problem, where you have a continuous value bounded by 0,1 with an infinite number of *p*, follows a **beta distribution**. Because the beta distribution is defined, you can instead just use that mathematical equation to come to the same conclusion as above. 

```{r}
#| label: q1-beta
#| echo: TRUE
#| eval: TRUE


ggplot() +
  stat_function(fun = function(x) dbeta(x, 4+1, 11+1), color = "black",
                size = 1) +
  theme_classic() + 
  xlab("p") + 
  ylab("")

```

#### Question 2: Using the posterior distribution from 1, compute the posterior predictive distribution for the next 5 tosses of the same globe. I recommend you use the sampling method.

```{r}
#| label: q2
#| echo: TRUE
#| eval: TRUE

# sample our posterior (beta distribution with 4 W, 11 L)
post_samples <- rbeta(1e4, 4+1, 11+1)

# sim globe function
sim_globe <- function(p, N){
  
  sample(c("W", "L"), size = N, prob=c(p, 1-p), replace = T)

}

# simulate posterior predictive distribution for next 5 tosses of the globe
# how many of the 5 tosses will = W? Tested 1e4 times
pred_post <- sapply(post_samples, function(p) sum(sim_globe(p, 5)=="W"))

# create table that counts number of incidences of each option (i.e., how many times will 0 W appear, 1 W, 2 W, 3 W, 4 W, 5 W?)
tab_post <- table(pred_post)
pred_post_df <- as.data.frame(tab_post)

ggplot(pred_post_df, aes(as.factor(pred_post), Freq)) + 
  geom_col() + 
  theme_classic() + 
  xlab("number of W") + 
  ylab("count")
```

#### Question 3: Use the posterior predictive distribution from 2 to calculate the probability of 3 or more water samples in the next 5 tosses.

#### Question 4: Suppose you observe W = 5 water points, but you forgot to write down how many times the globe was tossed, so you don't know the number of land pints L. Assume that *p* = 0.7 and compute the posterior distribution of the number of tosses *N*. Hint: Use the binomial distribution.