---
title: "Homework - Week 02"
author: "Isabella C. Richmond"
format: html
---

```{r, include = FALSE}
source('R/packages.R')
```

#### Question 1: From the Howell1 dataset, consider only the people younger than 13 years old. Estimate the causal association between age and weight. Assume that age influences weight through two paths. First, age influences height, and height influences weight. Second, age directly influences weight through age-related changes in muscle growth and body proportions.

Draw the DAG that represents these causal relationships. And then write
a generative simulation that takes age as an input and simulates height and
weight, obeying the relationships in the DAG.

DAG:
```{r}
# DAG
dag <- dagify(
    weight ~ age + height,
    height ~ age,
    exposure = 'age',
    outcome = 'weight'
)

ggdag(dag) +
    theme_dag()

```
Generative Simulation: 

```{r}
# A = age, H = height, W = weight
bAH <- 5 # as age increases, height increases by a lot
bHW <- 0.5 # as height increases, weight proportionally increases cm/kg
bAW <- 0.1 # as age increases, there is a small direct increase in weight

sim_AW <- function(A, bAH, bHW, bAW){
  N <- length(A)
  H <- rnorm(N, bAH*A, 2)
  W <- rnorm(N, bHW*H + bAW*A, 2)
  data.frame(A, H, W)
}

# randomly generated individuals 
A <- runif(n=20, min=1, max=13)

test_sim <- sim_AW(A, bAH, bHW, bAW)

# view relationship
ggplot(data = test_sim, aes(x=A, y=W)) + 
  geom_point() + 
  theme_classic()

```
Have some weird things happening here (decreases in weight with age) but overall a positive-ish relationship between age-weight that is partially dependent on height.


#### Question 2: Use a linear regression to estimate the total causal effect of each year of growth on weight.

Total causal effect means that we do not account for the effect of height. It is a linear regression of weight ~ age (where age is a proxy for year of growth).

```{r}
data(Howell1)
d <- Howell1 %>% 
  filter(age <= 13) %>%
  select(c(weight, age))

# get a list of priors present in the model
default_prior <- get_prior(weight ~ age, data = d, family = gaussian())

# get model
tar_load(h02_mAW)
mAW <- h02_mAW

# model formula
mAW$formula

# priors
mAW$prior

# check diagnostics
plot(mAW)

# get a summary of posterior distribution
posterior_summary(mAW)
```


#### Question 3: Now suppose the causal association between age and weight might be different for boys and girls. Use a single linear regression, with a categorical variable for sex, to estimate the total causal effect of age on weight separately for boys and girls. How do girls and boys differ? Provide one or more posterior contrasts as a summary. 

Updated DAG: 

```{r}
# DAG
dag <- dagify(
    weight ~ height + age + sex,
    height ~ age + sex,
    exposure = 'age',
    outcome = 'weight'
)

ggdag(dag) +
    theme_dag()

```

Model, stratified by sex:
```{r}
data(Howell1)
d2 <- Howell1 %>% 
  filter(age <= 13) %>%
  mutate(sex = as.factor(case_when(male == 0 ~ 1, 
                         male == 1 ~ 2))) %>%
  select(c(weight, age, sex))

# get a list of priors present in the model
default_prior2 <- get_prior(weight ~ age + sex, data = d2, family = gaussian())

# load model
tar_load(h02_mAWS)
mAWS <- h02_mAWS

# model formula
mAWS$formula

# priors
mAWS$prior

# check diagnostics
plot(mAWS)

# get a summary of posterior distribution
posterior_summary(mAWS)

# get prediction options 
pred_vals <- data.frame(age = seq(0, 13, 0.5))

# simulate from model where sex = 1 and sex = 2 
post_pred_female <- predicted_draws(mAWS, pred_vals %>% mutate(sex = 1)) 
post_pred_male <- predicted_draws(mAWS, pred_vals %>% mutate(sex = 2))

# prediction table 
post_differences <- data.frame(
  predict_sex_female = post_pred_female$.prediction,
  predict_sex_male = post_pred_male$.prediction,
  age = post_pred_female$age
) %>%
  mutate(diff = predict_sex_male - predict_sex_female)



ggplot(post_differences, aes(age, diff)) +
	stat_lineribbon() +
  scale_fill_manual(values = met.brewer("VanGogh3", 3)) +
	labs(
	  x = 'age',
		y = 'predicted difference in weight (males-females)') +
  theme_classic()
```

#### Question 4: The data in data(Oxboys) (rethinking package) are growth records for 26 boys measured over 9 periods. I want you to model their growth. Specifically, model the increments in growth from one period (Occasion in the data table) to the next. Each increment is simply the difference between height in one occasion and height in the previous occasion. Since none of these boys shrunk during the study, all of the growth increments are greater than zero. Estimate the posterior distribution of these increments. Constrain the distribution so it is always positive-it should not be possible for the model to think that boys can shrink from year to year. Finally compute the posterior distribution of the total growth over all 9 occasions.







