---
title: "Lecture 17 - Measurement & Misclassification"
author: Isabella C. Richmond
date: May 22, 2024
format: html
---

```{r, echo = F, warning = F, message = F}
source("R/packages.R")
```

## Rose / Thorn

*Rose:*

*Thorn:*

## Measurement Error

-   many variables are proxies of the cause of interest
    -   proxies are descendants of the things we wish we could measure
    -   often they are measured with error
-   don't consider how things are measured
-   measurement error can have many effects on estimates - reduce or increase effect estimates
-   no general rule about what measurement error will do to an estimate

## Modeling Measurement

-   divorce, marriage, and age statistics are measured with error and the amount of error varies by state

    -   imbalance in evidence quality

    -   potential confounding through measurement error because measurement error is associated with other variables (e.g., population size)

-   states with larger populations have less uncertainty/higher quality data

-   confounding because measurement is influenced by population size but then effects such as divorce rate can also be influenced by population size

```{r, eval = T, echo = F}
library(ggdag)
library(ggplot2)

dagified <- dagify(
    # outline relationships between your variables
    M_obs ~ e_M + M_act + A_act,
    A_obs ~ e_A + A_act,
    D_act ~ M_act + A_act,
    D_obs ~ D_act + e_D,
    e_M ~ P,
    e_A ~ P,
    e_D ~ P
    )
    

ggplot(dagified) +
    theme_dag()
```

-   think like a graph!

    -   how do I model the network of causes? ie full luxury bayes

-   In this dag there is a model for D_act, model for D_obs, model for M_act

-   D_act

    -   $D_i \sim Normal(\mu_i, \sigma)$

    -   $\mu_i = \alpha + \beta_A A_i + \beta_M M_i$

-   D_obs

    -   $D^*_i = D_i + e_{D,i}$

    -   $e_{D,i} \sim Normal(0, S_i)$

    -   $D^*_i \sim Normal(D_i, S_i)$

```{r, eval = F}
m15.1 <- ulam(
  alist(
    # model for D* (observed)
    D_obs ~ dnorm(D_true, D_sd),
    
    # model for D (unobserved)
    vector[N]:D_true ~ dnorm(mu, sigma),
    mu <- a + bA*A + bM*M,
    a ~ dnorm(0, 0.2),
    bA~ dnorm(0, 0.5),
    bM ~ dnorm(0,0,5), 
    sigma ~ dexp(1)
  ), data = dlist, chains = 4, cores = 4
)
```

-   can also add marriage rate in the same way

-   Full luxury bayes allows models to share information with each other

## Misclassification

-   categorical version of measurement error

-   if effect size is small, the effect may be on the same order of magnitude as the error rates

-   $X_i \sim Bernoulli(p_i)$

-   Generative model

    -   $X_i \sim Bernoulli(p_i)$

    -   $logit(p_i) = \alpha + \mu_{M[i]} + \delta_{D[i]}$

    -   M = mom, D = dyad

-   doing things on the log scale avoids underflowing to 0 or overflowing to 1 in probability space

    -   log is the solution to this

-   related models: hurdle models and occupancy models

-   related questions: rating and assessment

## Floating Point Monsters

-   changing probability statements to logarithms is what is done automatically in stan

-   when p is small, use convenience functions
